---
title: "Possum Regression"
author: "Guillermo Luigui Ubaldo Nieto Angarita, Daniel Hernando Moyano Salamanca, Jairo Antonio Viteri Rojas"
date: "2025-03-05"
output: html_document
---

# Taller Regresión Zarigüeya

Es fundamental que el proyecto considere, como mínimo, los siguientes tópicos:
- Objetivo del análisis estadístico del conjunto de datos
- Descripción del contexto al que pertenecen los datos
- Descripción y escala de medición de las variables consideradas
- El análisis descriptivo de los datos
- Estimación de modelos de regresión y su interpretación
- Comparación de modelos
- Validación de supuestos
- Análisis diagnóstico

## Paso 1: Contexto

## Paso 2: Problema


- Can we use total length to predict a possum's head length?
- Which possum body dimensions are most correlated with age and sex?
- Can we classify a possum's sex by its body dimensions and location?
- Can we predict a possum's trapping location from its body dimensions?


## Paso 3: Objetivos

## Paso 4: Instalar Liberias

## Psso 5: Cargar Liberias
```{r}
library(rlang)
library(ggplot2)
library(corrplot)
library(vcd)
library(broom)
library(effectsize)
library(nortest)
library(lmtest)
library(car)
library(dplyr)
library(tidyr)
library(patchwork)
library(reshape2)
```
## Paso 6: Importación de Datos y Lectura del Conjunto de Datos
### Paso 6.1: Asignar la ruta de los datos
```{r}
setwd("..")
path <- getwd()
file_path <- file.path(path, "data", "possum.csv")
print(file_path)
```


### Paso 6.2: Importar los datos 
```{r}
data <- read.csv(file_path)
```


## Paso 7: Analizando los Datos
### Paso 7.1: Inspección Básica de Datos
```{r}
print(dim(data))
```
```{r}
head(data)
```

```{r}
tail(data)
```

### Paso 7.2: Resumen Estadístico Básico
```{r}
summary(data)
```

## Paso 8: Gestión de Duplicados y Valores Faltantes
### Paso 8.1: Gestión de Duplicados
#### Paso 8.1.1: Identificación de Duplicados
```{r}
duplicates <- duplicated(data)
sum(duplicates)
```

## Paso 9: Reducción de Datos
```{r}
data <- data[, !(names(data) %in% c("case"))]
head(data)
```

## Paso 10: Limpieza y Manipulación de Datos
### Paso 10.1: minusculas

```{r}
data$Pop <- tolower(data$Pop)
```


## Paso 10.2: Identificar tipos de columnas

```{r}
# Identificar tipos de columnas
numeric_columns <- names(data)[sapply(data, is.numeric)]
categorical_columns <- names(data)[sapply(data, is.character) | sapply(data, is.factor)]

# Clasificar variables numéricas en discretas o continuas
discrete_columns <- names(data)[sapply(data, function(x) is.numeric(x) && all(x == floor(x), na.rm = TRUE))]
continuous_columns <- setdiff(numeric_columns, discrete_columns)

# Clasificar variables categóricas en nominales u ordinales
ordinal_columns <- names(data)[sapply(data, is.ordered)]
nominal_columns <- setdiff(categorical_columns, ordinal_columns)

target_column <- "hdlngth"
print("Variable Objetivo:")
print(target_column)

# Imprimir resultados
print("Numéricas Discretas:")
print(discrete_columns)

print("Numéricas Continuas:")
print(continuous_columns)

print("Categóricas Ordinales:")
print(ordinal_columns)

print("Categóricas Nominales:")
print(nominal_columns)
```


## Paso 9: Resumen Estadístico
### Paso 9.1: Resumen
```{r}
summary(data)
```


### Paso 9.2: Tendencia Central
```{r}
if (!require(moments)) install.packages("moments", dependencies = TRUE)
library(moments)

summary_stats <- function(column) {
  mean_val <- mean(column, na.rm = TRUE)
  median_val <- median(column, na.rm = TRUE)
  mode_val <- as.numeric(names(sort(table(column), decreasing = TRUE))[1])  # Moda
  
  std_dev <- sd(column, na.rm = TRUE)
  variance <- var(column, na.rm = TRUE)
  
  range_val <- max(column, na.rm = TRUE) - min(column, na.rm = TRUE)
  iqr_val <- IQR(column, na.rm = TRUE)
  
  skewness_val <- skewness(column, na.rm = TRUE)
  kurtosis_val <- kurtosis(column, na.rm = TRUE)
  
  return(c(mean = mean_val, median = median_val, mode = mode_val, 
           std_dev = std_dev, variance = variance, range = range_val, iqr = iqr_val,
           skewness = skewness_val, kurtosis = kurtosis_val))
}

numeric_columns <- data[sapply(data, is.numeric)]
result <- as.data.frame(t(sapply(numeric_columns, summary_stats)))

print(result)
```


## Paso 11: Análisis Univariado
### Paso 11.1: Análisis Univeriado de Variables Categorías
#### Paso 11.1.1: Análisis Univeriado de Variables Nominales

```{r}
for (column in categorical_columns) {
  p <- ggplot(data = data, aes(x = !!sym(column))) +
    geom_bar()
  print(p)
}
```

### Paso 11.2: Análisis Univeriado de Variables Númericas
#### Paso 11.2.1: Análisis Univeriado de Variables Discretas

```{r}
for (column in discrete_columns) {
  p <- ggplot(data = data, aes(x = !!sym(column))) +
    geom_bar()
  print(p)
}
```


#### Paso 11.2.2: Análisis Univeriado de Variables Continuas

```{r}
for (column in continuous_columns) {
  p <- ggplot(data = data, aes(x = !!sym(column))) + 
    geom_histogram()
  print(p)
}
```


## Paso 12: Análisis Bivariado
### Paso 12.1: Análisis Bivariado de Variables Categorías
#### Paso 12.1.1: Análisis Bivariado de Variables Nominales
##### Paso 12.1.1.1: Box Plots

```{r}
for (column in categorical_columns) {
  p <- ggplot(data = data, aes(x = !!sym(column), y = !!sym(target_column))) + 
    geom_boxplot()
  print(p)
}
```

##### Paso 12.1.1.1: Kernel Density plots 

```{r}
for (column in categorical_columns) {
  p <- ggplot(data = data, aes(x = !!sym(target_column), fill = !!sym(column))) + 
    geom_density(alpha = 0.4)
  print(p)
}
```


### Paso 12.2: Análisis Bivariado de Variables Númericas
#### Paso 12.2.1: Análisis Univeriado de Variables Discretas
##### Paso 12.2.1.1: Box Plots

```{r}
for (column in discrete_columns) {
  p <- ggplot(data = data, aes(x = factor(!!sym(column)), y = !!sym(target_column))) + 
    geom_boxplot() +
    xlab(column) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(p)
}
```


##### Paso 12.2.1.1: Kernel Density plots

```{r}
for (column in discrete_columns) {
  p <- ggplot(data = data, aes(x = !!sym(target_column), fill = factor(!!sym(column)))) + 
    geom_density(alpha = 0.4)
  print(p)
}
```


#### Paso 12.2.2: Análisis Univeriado de Variables Continuas

```{r}
for (column in continuous_columns) {
  if (column != target_column) {
    p <- ggplot(data = data, aes(x = !!sym(column), y = !!sym(target_column))) + 
      geom_point()
    print(p)
  }
}
```


## Paso 13: Multivariate Analysis

```{r}
correlation_matrix <- cor(data[, names(numeric_columns)], use = "pairwise.complete.obs")
print(correlation_matrix)

```

```{r}
corrplot(correlation_matrix, method = "color", tl.cex = 0.8)
```


```{r}
library(dplyr)

# Create a pairwise chi-square test for all categorical variables
chi_square_results <- expand.grid(categorical_columns, categorical_columns) %>%
  filter(Var1 != Var2) %>%
  mutate(chi2_p_value = mapply(function(x, y) {
    table_data <- table(data[[x]], data[[y]])
    chisq.test(table_data)$p.value
  }, Var1, Var2))

print(chi_square_results)

```


```{r}


for (i in 1:(length(categorical_columns)-1)) {
  for (j in (i+1):length(categorical_columns)) {
    formula <- as.formula(paste("~", categorical_columns[i], "+", categorical_columns[j]))
    mosaic(formula, data = data, shade = TRUE, legend = TRUE)
  }
}

```

```{r}
# Example: Assume numeric_columns contains the names of numeric variables
numeric_data <- data[names(numeric_columns)]

# Compute p-values using nested sapply() with proper indexing
cor_pvalues <- sapply(seq_along(numeric_columns), function(i) {
  sapply(seq_along(numeric_columns), function(j) {
    if (i != j) {
      cor.test(numeric_data[[i]], numeric_data[[j]])$p.value
    } else {
      NA
    }
  })
})

# Convert to matrix and set row/column names
cor_pvalues <- as.matrix(cor_pvalues)
rownames(cor_pvalues) <- colnames(cor_pvalues) <- names(numeric_columns)

# Print the matrix of p-values
print(cor_pvalues)
```
```{r}

# Melt the matrix into long format
pvalue_melted <- melt(cor_pvalues, na.rm = TRUE)

# Add a column for significance (e.g., p < 0.05)
pvalue_melted$significant <- ifelse(pvalue_melted$value < 0.05, "Yes", "No")

# Optional: Add asterisks for significance levels
pvalue_melted$label <- ifelse(
  pvalue_melted$value < 0.001, "***",
  ifelse(pvalue_melted$value < 0.01, "**",
         ifelse(pvalue_melted$value < 0.05, "*", ""))
)
```


```{r}
ggplot(pvalue_melted, aes(x = Var1, y = Var2, fill = significant)) +
  geom_tile(color = "white") +
  geom_text(aes(label = label), color = "black", size = 4) +  # Add asterisks
  scale_fill_manual(values = c("Yes" = "#FF6B6B", "No" = "#C4DDFF")) +
  labs(
    title = "Statistical Significance of Pairwise Correlations",
    x = "", y = "", fill = "Significant (p < 0.05)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```

```{r}
# Extract numeric variables (excluding 'hdlngth' to avoid self-correlation)
numeric_vars <- names(data)[sapply(data, is.numeric)]
numeric_vars <- numeric_vars[numeric_vars != "hdlngth"]  # Remove target

# Compute correlations and p-values between 'hdlngth' and all other numeric vars
results <- sapply(numeric_vars, function(var) {
  test <- cor.test(data$hdlngth, data[[var]])
  c(correlation = test$estimate, pvalue = test$p.value)
})

# Transpose and convert to dataframe
results_df <- as.data.frame(t(results))
results_df$variable <- rownames(results_df)
```


```{r}
# Step 1: Compute correlations with hdlngth
numeric_vars <- names(data)[sapply(data, is.numeric)]
numeric_vars <- numeric_vars[numeric_vars != "hdlngth"]  # Exclude target

results <- lapply(numeric_vars, function(var) {
  test <- cor.test(data$hdlngth, data[[var]])
  data.frame(
    variable = var,
    cor = test$estimate,
    pvalue = test$p.value
  )
})

numeric_results_df <- do.call(rbind, results)

# Step 2: Create the plot (fixed version)
ggplot(numeric_results_df, aes(x = reorder(variable, cor), y = cor, 
                      fill = pvalue < 0.05)) +
  geom_col() +  # Use geom_col() instead of geom_bar(stat = "identity")
  geom_text(aes(label = ifelse(pvalue < 0.05, "*", "")), 
            vjust = 0.5, hjust = ifelse(numeric_results_df$cor > 0, -0.2, 1.2),
            size = 5, color = "black") +
  scale_fill_manual(values = c("TRUE" = "#FF6B6B", "FALSE" = "#C4DDFF"),
                    name = "Significant\n(p < 0.05)") +
  labs(
    title = "Correlation with hdlngth",
    x = "Variable",
    y = "Pearson's r"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```


```{r}
# Step 0: Verify hdlngth exists
if(!"hdlngth" %in% names(data)) {
  stop("Error: 'hdlngth' column not found in the dataset")
}

# Step 1: Identify categorical variables (factors/characters with <10 unique values)
cat_vars <- names(data)[sapply(data, function(x) {
  (is.factor(x) | is.character(x)) & n_distinct(x, na.rm = TRUE) < 10
})]

# Remove NA-only variables
cat_vars <- cat_vars[sapply(cat_vars, function(x) !all(is.na(data[[x]])))]

# Step 2: Statistical tests with error handling


results <- lapply(cat_vars, function(var) {
  tryCatch({
    if (n_distinct(data[[var]], na.rm = TRUE) == 2) {
      # t-test for binary variables
      test <- t.test(data$hdlngth ~ data[[var]])
      data.frame(
        variable = var,
        method = "t-test",
        p.value = test$p.value,
        effect_size = cohens_d(data$hdlngth ~ data[[var]])$Cohens_d
      )
    } else {
      # ANOVA for 3+ categories
      model <- aov(data$hdlngth ~ factor(data[[var]]))
      p.value <- summary(model)[[1]]$"Pr(>F)"[1]
      data.frame(
        variable = var,
        method = "ANOVA",
        p.value = p.value,
        effect_size = eta_squared(model)$Eta2
      )
    }
  }, error = function(e) {
    data.frame(
      variable = var,
      method = NA_character_,
      p.value = NA_real_,
      effect_size = NA_real_
    )
  })
})

# Combine results and remove failed tests
results_df <- do.call(rbind, results) %>% 
  filter(!is.na(method))
```
```{r}
library(ggplot2)
library(dplyr)

# Plot 1: Statistical significance
ggplot(results_df, aes(x = reorder(variable, -p.value), 
                     y = -log10(p.value), 
                     fill = p.value < 0.05)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  geom_text(aes(label = ifelse(p.value < 0.05, "*", "")), 
            size = 6, vjust = 0.7, hjust = -0.2) +
  scale_fill_manual(values = c("TRUE" = "#ff6b6b", "FALSE" = "#7fcdbb")) +
  labs(title = "Relationship Between hdlngth and Categorical Variables",
       subtitle = "Dashed line shows p = 0.05 significance threshold",
       x = NULL,
       y = "-log10(p-value)") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")

# Plot 2: Effect sizes
ggplot(results_df, aes(x = reorder(variable, abs(effect_size)), 
                     y = effect_size, 
                     fill = p.value < 0.05)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, linetype = "solid") +
  scale_fill_manual(values = c("TRUE" = "#ff6b6b", "FALSE" = "#7fcdbb")) +
  labs(title = "Effect Sizes for Categorical Variables",
       x = NULL,
       y = ifelse(all(results_df$method == "t-test"), 
                  "Cohen's d", 
                  "Eta-squared")) +
  coord_flip() +
  theme_minimal(base_size = 14)
```

## Paso 14: Pruebas de hipótesis



## Paso 15: Modelamiento
### Paso 15.1: Entrenamiento

### Paso 15.2: Pruebas de supuestos
#### Paso 15.2.1: Supuestos de regresion lineal
https://rpubs.com/acastro/regresion_lineal_supuestos

1. Linealidad
2. Normalidad
3. Varianza constante (Homocedasticidad)
4. Independencia



```{r}
linear_model1 <- lm(hdlngth ~ ., data = data)
summary(linear_model1)
```


```{r}
linear_model2 <- lm(hdlngth ~ age + skullw + totlngth + taill + footlgth + eye + chest + belly + Pop, data = data)
summary(linear_model2)
```


```{r}
linear_model3 <- lm(hdlngth ~ skullw + totlngth + taill + footlgth + Pop, data = data)
summary(linear_model3)
```
```{r}
linear_model4 <- lm(hdlngth ~ skullw + totlngth, data = data)
summary(linear_model4)
```


#### Paso 15.2.2: Prueba de Linealidad

```{r}
# Create combined data frame
residual_data <- lapply(1:4, function(i) {
  model <- get(paste0("linear_model", i))
  data.frame(
    Model = paste("Model", i),
    Fitted = model$fitted.values,
    Residuals = residuals(model)
  )
}) %>% bind_rows()

# Create plot
ggplot(residual_data, aes(x = Fitted, y = Residuals, color = Model)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~Model, ncol = 2) +
  labs(title = "Residuals vs Fitted Values Comparison",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal() +
  theme(legend.position = "none")
```


#### Paso 15.2.3: Prueba de Normalidad

```{r}
# Create empty list to store results
normality_results <- list()

# Run Lilliefors test for each model
for (i in 1:4) {
  model <- get(paste0("linear_model", i))
  e <- residuals(model)
  normality_results[[i]] <- lillie.test(e)
  
  cat("\n--- Model", i, "---\n")
  print(normality_results[[i]])
  


# Create summary table
normality_summary <- lapply(1:4, function(i) {
  model <- get(paste0("linear_model", i))
  test_result <- lillie.test(residuals(model))
  data.frame(
    Model = paste("Model", i),
    Statistic = test_result$statistic,
    p.value = test_result$p.value,
    Normality = ifelse(test_result$p.value > 0.05, "Normal", "Non-normal")
  )
}) %>% bind_rows()

# Print formatted table
print(normality_summary)
}
```

```{r}
# Create list of plots
plot_list <- lapply(1:4, function(i) {
  model <- get(paste0("linear_model", i))
  ggplot(data.frame(res = residuals(model)), 
         aes(sample = res)) +
    stat_qq() +
    stat_qq_line(color = "red") +
    labs(title = paste("Model", i), 
         x = "Theoretical Quantiles",
         y = "Sample Quantiles")
})

# Combine plots
(plot_list[[1]] | plot_list[[2]]) / 
(plot_list[[3]] | plot_list[[4]])
```

```{r}
library(car)

# Set up 2x2 plot layout
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))  # Adjust margins

# Create QQ-plots for each model
for (i in 1:4) {
  model <- get(paste0("linear_model", i))
  qqPlot(residuals(model), 
         main = paste("Model", i, "QQ Plot"),
         ylab = "Residuals",
         envelope = 0.95,  # 95% confidence band
         col.lines = "red",
         id = FALSE)  # Turn off point labeling
}

# Reset layout
par(mfrow = c(1, 1))
```



#### Paso 15.2.4: Prueba de Varianza Constante
##### Paso 15.2.4.1: Goldfeld-Quandt Test

```{r}
# Create empty list to store results
gq_results <- list()

# Run Goldfeld-Quandt test for each model
for (i in 1:4) {
  model <- get(paste0("linear_model", i))
  gq_results[[i]] <- gqtest(model)
  
  cat("\n--- Model", i, "---\n")
  print(gq_results[[i]])
}
```


```{r}
plot_list <- lapply(1:4, function(i) {
  model <- get(paste0("linear_model", i))
  ggplot(data.frame(Fitted = fitted(model),
                   Residuals = residuals(model)),
         aes(x = Fitted, y = Residuals)) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "loess", color = "red", se = FALSE) +
    labs(title = paste("Model", i),
         x = "Fitted Values",
         y = "Residuals") +
    theme_minimal()
})

# Combine plots
(plot_list[[1]] | plot_list[[2]]) / 
(plot_list[[3]] | plot_list[[4]])
```

##### Paso 15.2.4.2: Breusch-Pagan Test

```{r}
for (i in 1:4) {
  model <- get(paste0("linear_model", i))
  bp_test <- bptest(model)
  
  cat("\n--- Model", i, "---\n")
  print(bp_test)
}
```

```{r}
plot_list <- lapply(1:4, function(i) {
  model <- get(paste0("linear_model", i))
  ggplot(data.frame(Fitted = fitted(model),
                   Residuals = residuals(model)),
         aes(Fitted, Residuals)) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "loess", color = "red", se = FALSE) +
    labs(title = paste("Model", i, 
                      ifelse(bptest(model)$p.value < 0.05, 
                             "(Heteroskedastic)", 
                             "(Homoskedastic)")),
         x = "Fitted Values",
         y = "Residuals") +
    theme_minimal()
})

# Combine plots
(plot_list[[1]] | plot_list[[2]]) / 
(plot_list[[3]] | plot_list[[4]])
```



### Paso 15.2.5: Prueba de Independencia

```{r}
# Create empty list to store results
dw_results <- list()

# Run Durbin-Watson test for each model
for (i in 1:4) {
  model <- get(paste0("linear_model", i))
  dw_results[[i]] <- durbinWatsonTest(model)
  
  cat("\n--- Model", i, "---\n")
  print(dw_results[[i]])
}
```

```{r}
plot_list <- lapply(1:4, function(i) {
  model <- get(paste0("linear_model", i))
  ggplot(data.frame(Order = 1:length(residuals(model)),
                   Residuals = residuals(model)),
         aes(x = Order, y = Residuals)) +
    geom_point() +
    geom_line(alpha = 0.5) +
    geom_smooth(method = "loess", color = "red", se = FALSE) +
    labs(title = paste("Model", i, "Residuals"),
         x = "Observation Order",
         y = "Residuals") +
    theme_minimal()
})

# Combine plots
(plot_list[[1]] | plot_list[[2]]) / 
(plot_list[[3]] | plot_list[[4]])
```



### Paso 15.3: Evaluación
1. MSE
2. MAD
3. MAPE
4. R2
5. AIC
6. BIC



